Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2018-01-16 manuscript/manuscript.tex
Bookdown

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[essd, classical]{copernicus}

%% Journal abbreviations (please use the same for discussion papers and final revised papers)

% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

%% My packages
\usepackage{amssymb} %symbole de maths
\usepackage{amsmath} %idem
\usepackage{longtable}
\setlength{\LTleft}{-5cm plus 1 fill}
\setlength{\LTright}{-5cm plus 1 fill}
%% New commands
\newcommand{\logit}{\text{logit}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\R}{\textnormal{\sffamily{\bfseries{R}}}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

\begin{document}

\title{Combining global tree cover loss data with historical national
  forest-cover maps to look at six decades of deforestation and forest
  fragmentation in Madagascar}

% \Author[affil]{given_name}{surname}

\Author[1, 2, 3]{Ghislain}{Vieilledent}
\Author[4]{Clovis}{Grinand}
\Author[4]{Fety A.}{Rakotomalala}
\Author[5]{Rija}{Ranaivosoa}
\Author[5]{Jean-Roger}{Rakotoarijaona}
\Author[6, 7]{Thomas F.}{Allnutt}
\Author[1]{Frédéric}{Achard}

\affil[1]{Joint Research Center of the European Commission, Bio-economy Unit, I-21027 Ispra (VA), ITALY}
\affil[2]{Cirad, UPR Forêts et Sociétés, F-34398 Montpellier, FRANCE}
\affil[3]{Forêts et Sociétés, Univ. Montpellier, Cirad, Montpellier, FRANCE}
\affil[4]{ETC Terra, F-75020 Paris, FRANCE}
\affil[5]{Office National pour l'Environnement, 101 Antananarivo, MADAGASCAR}
\affil[6]{Wildlife Conservation Society, 101 Antananarivo, MADAGASCAR}
\affil[7]{GreenInfo Network, Oakland, California, USA}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

\runningtitle{Six decades of deforestation in Madagascar}

\runningauthor{G. Vieilledent}

\correspondence{Ghislain Vieilledent (ghislain.vieilledent@cirad.fr)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}
\maketitle

\begin{abstract}

  The island of Madagascar has an unparalleled biodiversity, mainly
  located in the tropical forests of the island, which is highly
  threatened by anthropogenic deforestation. Scattered forest maps
  from past studies at national level with substantial gaps (due to
  presence of cloud cover on satellite imagery) prevent the analysis
  of long-term deforestation trends in Madagascar. In this study, we
  propose a new approach combining historical (1953-2000) national
  forest-cover maps with recent (2001-2014) global annual tree cover
  loss data to look at six decades (1953-2014) of deforestation and
  forest fragmentation in Madagascar. We produced new forest-cover
  maps at 30~m resolution over the full territory of Madagascar for
  the year 1990, and annually from 2000 to 2014. We estimated that
  Madagascar has lost 44\% of its natural forest cover over the period
  1953-2014 (including 37\% over the period 1973-2014). Natural
  forests cover 8.9 Mha in 2014 (15\% of the national territory) which
  are divided into 4.4 Mha (50\%) of moist forests, 2.6 Mha (29\%) of
  dry forests, 1.7 Mha of spiny forests (19\%) and 177,000 ha (2\%) of
  mangroves. Since 2005, the annual deforestation rate has
  progressively increased in Madagascar to reach 99,000 ha/yr during
  2010-2014 (corresponding to a rate of 1.1\%/yr). This increase is
  probably due to rapid population growth (close to 3\%/yr) and to
  poor law enforcement in the country. Around half of the forest
  (46\%) is now located at less than 100~m from the forest
  edge. Accurate forest-cover change maps can be used to assess the
  effectiveness of past and current conservation programs and
  implement new strategies for the future. In particular, forest maps
  and estimates can be used in the REDD+ framework which aims at
  ``Reducing Emissions from Deforestation and Forest Degradation'' and
  for optimizing the current protected area network.

\end{abstract}

\copyrightstatement{The article is distributed under the Creative Commons Attribution 4.0 License.}

\section{Introduction}
\label{introduction}

Separated from the African continent and the Indian plate about 165
and 88 million years ago respectively \citep{Ali2008}, the flora and
fauna of Madagascar followed its own evolutionary path. Isolation
combined with a high number of micro-habitats \citep{Pearson2009} has
led to Madagascar's exceptional biodiversity both in term of number of
species and endemism in many taxonomic groups \citep{Crottini2012,
  Goodman2005}. Most of the biodiversity in Madagascar is
concentrated in the tropical forests of the island which can be
divided into four types: the moist forest in the East, the dry forest
in the West, the spiny forest in the South and the mangroves on the
West coast \citep{Vieilledent2016}. This unparalleled biodiversity is
severely threatened by deforestation \citep{Harper2007,
  Vieilledent2013} associated with human activities such as slash-and-burn
agriculture and pasture \citep{Scales2011}. Tropical forests in
Madagascar also store a large amount of carbon \citep{Vieilledent2016}
and high rates of deforestation in Madagascar are responsible for
large CO$_2$ emissions in the atmosphere
\citep{Achard2014}. Deforestation threatens species survival by
directly reducing their available habitat \citep{Brooks2002,
  Tidd2001}. Forest fragmentation can also lead to species extinction
by isolating populations from each other and creating forest patches
too small to maintain viable populations
\citep{Saunders1991}. Fragmentation also increases forest edge where
ecological conditions (such as air temperature, light intensity and
air moisture) can be dramatically modified, with consequences on the
abundance and distribution of species \citep{Murcia1995}. Forest
fragmentation can also have substantial effects on forest carbon
storage capacity, as carbon stocks are much lower at the forest edge
than under a closed canopy \citep{Brinck2017}. Moreover, forest
carbon stocks vary spatially due to climate or soil factors
\citep{Saatchi2011, Vieilledent2016}. As a consequence, accurate and
spatially explicit maps of forest-cover and forest-cover change are
necessary to monitor biodiversity loss and carbon emissions from
deforestation and forest fragmentation, assess the efficiency of
present conservation strategies \citep{Eklund2016}, and implement new
strategies for the future \citep{Vieilledent2013,
  Vieilledent2016}. Simple time-series of forest-cover estimates, such
as those provided by the FAO Forest Resource Assessment report
\citep{Keenan2015} are not sufficient.

Unfortunately, accurate and exhaustive forest-cover maps are not
available for Madagascar for the last fifteen years (2000-2015).
\citet{Harper2007} produced maps of forest cover and forest cover
changes over Madagascar for the years \emph{c.}~1953, \emph{c.}~1973,
1990 and 2000. The \emph{c.}~1953 forest map was derived from the
visual interpretation of aerial photography at coarse scale
(1/1,000,000). Forest maps for the years \emph{c.}~1973, 1990, and
2000 were obtained from supervised classification of Landsat satellite
images at 60~m resolution (for the year 1973) or 30~m resolution (for
years 1990 and 2000) and can be used to derive more accurate estimates
of forest cover (89.5\% accuracy reported for the forest/non-forest
map of year 2000). Nonetheless, maps provided by \citet{Harper2007}
are not exhaustive (due to the presence of clouds in the satellite
imagery), e.g.~11 244~km2 are mapped as unknown cover type for the
year 2000. Using a similar supervised classification approach as in
\citet{Harper2007}, more recent maps have been produced for the
periods 2000-2005-2010 by national institutions, with the technical
support of international environmental NGOs \citep{MEFT2009,
  ONE2013}. Another set of recent forest-cover maps using an advanced
statistical tool for classification, the Random Forest classifier
\citep{Grinand2013, Rakotomalala2015}, was produced for the periods
2005-2010-2013 \citep{ONE2015}. However, these maps are either too old
to give recent estimates of deforestation \citep{MEFT2009, ONE2013},
include large areas of missing information due to images with high
percentage of cloud cover \citep{ONE2013}, or show large
mis-classification in specific areas, especially in the dry and spiny
forest domain for which the spectral answer has a strong seasonal
behavior due to the deciduousness of such forests (overall accuracy is
lower than 0.8 for the dry and spiny forests for the maps produced by
\citet{ONE2015}). Moreover, the production of such forest maps from a
supervised classification approach requires significant resources,
especially regarding the image selection step (required to minimize
cloud cover) and the training step (visual interpretation of a large
number of polygons needed to train the classification algorithm)
\citep{Rakotomalala2015}. Most of this work of image selection and
visual interpretation would need to be repeated to produce new forest
maps in the future using a similar approach.

Global forest or tree cover products have also been published recently
and can be tested at the national scale for
Madagascar. \citet{Kim2014} produced a global forest-cover change map
from 1990 to 2000 (derived from Landsat imagery). This product was
updated to cover the period 1975-2005
(\url{http://glcf.umd.edu/data/landsatFCC/}) but forest-cover maps
after 2005 were not produced. Moreover, the approach used in
\citet{Kim2014} did not accurately map the forests in the dry and
spiny ecosystems of Madagascar (see Fig. 8 in \citet{Kim2014}).
\citet{Hansen2013} mapped tree cover percentage, annual tree cover
loss and gain from 2000 to 2012 at global scale at 30 m
resolution. This product has since been updated and is now available
up to the year 2014 \citep{Hansen2013}. To map forest cover from the
\citet{Hansen2013} product, a tree cover threshold must be selected
(that defines forest cover). Selecting such a threshold is not
straightforward as the accuracy of the global tree cover map strongly
varies between forest types, and is substantially lower for dry
forests than for moist forests \citep{Bastin2017}. Moreover, the
\citet{Hansen2013} product does not provide information on
land-use. In particular the global tree cover map does not separate
tree plantations such as oil palm or eucalyptus plantations from
natural forests \citep{Tropek2014}. Thus, the global tree cover map
from \citet{Hansen2013} cannot be used alone to produce a map of
forest cover \citep{Tyukavina2017}.

In this study, we present a simple approach which combines the maps
from \citet{Harper2007} and products from \citet{Hansen2013} to derive
annual wall-to-wall forest-cover change maps over the period 2000-2014
for Madagascar. We use the forest-cover map provided by
\citet{Harper2007} for the year 2000 (defining the land-use) with the
tree cover loss product provided by \citet{Hansen2013} that we apply
only inside forest areas identified by \citet{Harper2007}. Similar to
the approach of \citet{Harper2007}, we also assess trends in
deforestation rates and forest fragmentation from \emph{c.} 1953 to 2014. The
approach described in this study can help assess the effectiveness of
current conservation strategies, and assist the implementation of
future strategies. Our approach could be easily extended to other
tropical countries that have at least one forest-cover map between
2000 and 2014. This approach can easily be repeated in the future when
tree cover loss products will be updated.

\section{Materials and Methods}
\label{materials-and-methods}

\subsection{Creation of new forest-cover maps of Madagascar from
1953 to 2014}

We produced annual forest/non-forest maps at 30~m resolution for the
full territory of Madagascar for the period 2000-2014 by combining the
forest map of year 2000 from \citet{Harper2007}, and the tree cover
percentage and annual tree cover loss maps over the period 2000-2014
from \citet{Hansen2013}. The 2000 Harper's forest map includes 208,000
ha of unclassified areas due to the presence of clouds on satellite
images, mostly (88\%) within the moist forest domain which covered
4.17 Mha in total in 2000. To provide a label (forest or non-forest)
to these unclassified pixels, we used the 2000 tree cover percentage
map of \citet{Hansen2013} by selecting a threshold of 75\% tree cover
to define forest cover as recommended by other studies for the moist
domain \citep{Achard2014, Aleman2017}. To do so, the Hansen's 2000
tree cover map was resampled on the same grid as the original Harper's
map at 30~m resolution using a bilinear interpolation. We thus obtained
a forest-cover map for the year 2000 covering the full territory of
Madagascar. We then combined this forest-cover map of the year 2000
with the annual tree cover loss maps from 2001 to 2014 provided by
\citet{Hansen2013} to create annual forest-cover maps from 2001 to
2014 at 30~m resolution. To do so, Hansen's tree cover loss maps were
resampled on the same grid as the original Harper's map at 30~m
resolution using a nearest-neighbor interpolation. We also completed
the Harper's forest map of year 1990 by filling unclassified areas
(due to the presence of clouds on satellite images) using our
forest-cover map of year 2000. To do so, we assumed that if forest was
present in 2000, the pixel was also forested in 1990. The remaining
unclassified pixels were limited to a relatively small total area of
\emph{c.} 8,000 ha. We labeled these residual pixels as non-forest, as
for the year 2000. Similarly we completed the Harper's forest map of
year 1973 by filling unclassified areas using our forest-cover map of
the year 1990 assuming that if forest was present in 1990, it was also
present in 1973. Contrary to the year 1990, the remaining unclassified
pixels for year 1973 corresponded to a significant total area of 3.3
million ha. We also reprojected the forest-cover map of year 1953 to a
common projection in order to compare the forest-cover area in 1953
with forest-cover areas at the following dates. This map was produced
by scanning a paper map derived from aerial photos, and thus could not
be perfectly aligned with the other maps produced through digital
processing of satellite imagery \citep{Harper2007}. Finally for all
forest-cover maps from 1973, the isolated single non-forest pixels
(i.e.~fully surrounded by forest pixels) were removed, assuming that
single non-forest pixels inside a forest patch were not corresponding
to deforestation (they might correspond to selective logging
activities). This allowed us to avoid counting very small scale events
(\textless 0.1 ha, such as selective logging) as forest
fragmentation. All the resulting maps are freely available at
\url{https://bioscenemada.cirad.fr/forestmaps}.

\subsection{Computing forest-cover areas and deforestation
rates}

From these new forest-cover maps, we calculated the total forest-cover
area for seven available years (1953-1973-1990-2000-2005-2010-2014),
and the annual deforested area and annual deforestation rate for the
corresponding six time periods between 1953 and 2014. The annual
deforestation rates were calculated using Eq.~\ref{eq:theta}
\citep{Puyravaud2003, Vieilledent2013}:

\begin{equation}
  \label{eq:theta}
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}
\end{equation}

In Eq.~\ref{eq:theta}, $\theta$ is the annual deforestation rate (in \%/yr),
$F_{t_2}$ and $F_{t_1}$ are the forest cover free of clouds at both
dates $t_2$ and $t_1$, and $t_2-t_1$ is the time-interval (in
years) between the two dates.

Because of the large unclassified area (3.3 million ha) in 1973, the
annual deforestation areas and rates for the two periods 1953-1973 and
1973-1990 are only partial estimates computed on the basis of the
available forest extent. Area and rate estimates are produced at the
national scale and for the four forest ecosystems present in
Madagascar: moist forest in the East, dry forest in the West, spiny
forest in the South, and mangroves on the Western coast
(Fig.~\ref{fig:ecoregion}). To define the forest domains, we used a
map from the MEFT (\emph{``Ministère de l'Environnement et des Forêts
  à Madagascar''}) with the boundaries of the four ecoregions in
Madagascar. Ecoregions were defined on the basis of climatic and
vegetation criteria using the climate classification by
\citet{Cornet1974} and the vegetation classification from the 1996
IEFN national forest inventory \citep{IEFN1996}. Because mangrove
forests are highly dynamic ecosystems that can expand or contract on
decadal scales depending on changes in environmental factors
\citep{Armitage2015}, a fixed delimitation of the mangrove ecoregion
on six decades might not be fully appropriate. As a consequence, our
estimates of the forest-cover and deforestation rates for mangroves in
Madagascar must be considered with this limitation.

\begin{figure}[h]
  \centering

    \includegraphics[width=8.3cm]{ecoregion.png}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Combining global tree cover loss data with historical national
forest-cover maps to look at six decades of deforestation and forest
fragmentation in Madagascar}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

\textbf{Ghislain Vieilledent\(^{1,2,\star}\), Clovis Grinand\(^{3}\),
Fety A. Rakotomalala\(^{3}\), Rija Ranaivosoa\(^{4}\), Jean-Roger
Rakotoarijaona\(^{4}\), Thomas F. Allnutt\(^{5,6}\), and Frédéric
Achard\(^{1}\)}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Joint Research Center of the European Commission, Bio-economy Unit
  (JRC.D.1), I-21027 Ispra (VA), ITALY
\item
  Cirad, UPR Forêts et Sociétés, F-34398 Montpellier, FRANCE
\item
  ETC Terra, F-75016 Paris, FRANCE
\item
  Office National pour l'Environnement, 101 Antananarivo, MADAGASCAR
\item
  Wildlife Conservation Society, 101 Antananarivo, MADAGASCAR
\item
  GreenInfo Network, Oakland, California, USA
\end{enumerate}

(*) /Email:
\href{mailto:ghislain.vieilledent@cirad.fr}{\nolinkurl{ghislain.vieilledent@cirad.fr}},
/Phone: +39.329.457.2273

\emph{Running headline:} Six decades of deforestation in Madagascar

\hypertarget{highlights}{%
\section{Highlights}\label{highlights}}

\begin{itemize}
\tightlist
\item
  We produced new 30m-resolution annual forest-cover maps for Madagascar
  for the period 2000-2014.
\item
  Madagascar has lost 44\% of its natural forest-cover over the period
  1953-2014.
\item
  Half of the tropical forest in Madagascar is now located at less than
  100m from forest edge.
\item
  Annual deforestation rate has increased in Madagascar since 2005 to
  reach 1.08\%/yr.
\item
  Conservation and development efforts must be intensified to save
  Madagascar forest and biodiversity.
\end{itemize}

\hypertarget{summary}{%
\section{Summary}\label{summary}}

The island of Madagascar has a unique biodiversity, mainly located in
the tropical forests of the island. This biodiversity is highly
threatened by anthropogenic deforestation. Existing historical forest
maps at national level are scattered and have substantial gaps which
prevent an exhaustive assessment of long-term deforestation trends in
Madagascar. In this study, we combine historical national forest cover
maps (covering the period 1953-2000) with a recent global annual tree
cover loss dataset (2001-2014) to look at six decades of deforestation
and forest fragmentation in Madagascar (from 1953 to 2014). We produced
new forest cover maps at 30~m resolution for the year 1990 and annually
from 2000 to 2014 over the full territory of Madagascar. We estimated
that Madagascar has lost 44\% of its natural forest cover over the
period 1953-2014 (including 37\% over the period 1973-2014). Natural
forests cover 8.9 Mha in 2014 (15\% of the national territory) and
include 4.4 Mha (50\%) of moist forests, 2.6 Mha (29\%) of dry forests,
1.7 Mha of spiny forests (19\%) and 177,000 ha (2\%) of mangroves. Since
2005, the annual deforestation rate has progressively increased in
Madagascar to reach 99,000 ha/yr during 2010-2014 (corresponding to a
rate of 1.1\%/yr). This increase is probably due to rapid population
growth (close to 3\%/yr) and to poor law enforcement in the country.
Around half of the forest (46\%) is now located at less than 100~m from
the forest edge. Accurate forest cover change maps can be used to assess
the effectiveness of past and current conservation programs and
implement new strategies for the future. In particular, forest maps and
estimates can be used in the REDD+ framework which aims at ``Reducing
Emissions from Deforestation and Forest Degradation'' and for optimizing
the current protected area network.

\textbf{Keywords}: biodiversity, climate-change, deforestation,
forest-fragmentation, Madagascar, tropical forest.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Separated from the African continent and the Indian plate about 165 and
88 million years ago respectively (Ali \& Aitchison
\protect\hyperlink{ref-Ali2008}{2008}), the flora and fauna of
Madagascar followed its own evolutionary path. Isolation combined with a
high number of micro-habitats (Pearson \& Raxworthy
\protect\hyperlink{ref-Pearson2009}{2009}) has led to Madagascar's
exceptional biodiversity both in term of number of species and endemism
in many taxonomic groups (Goodman \& Benstead
\protect\hyperlink{ref-Goodman2005}{2005}; Crottini \emph{et al.}
\protect\hyperlink{ref-Crottini2012}{2012}). Most of the biodiversity in
Madagascar is concentrated in the tropical forests of the island which
can be divided into four types: the moist forest in the East, the dry
forest in the West, the spiny forest in the South and the mangroves on
the West coast (Vieilledent \emph{et al.}
\protect\hyperlink{ref-Vieilledent2016}{2016}). This unparalleled
biodiversity is severely threatened by deforestation (Harper \emph{et
al.} \protect\hyperlink{ref-Harper2007}{2007}; Vieilledent, Grinand \&
Vaudry \protect\hyperlink{ref-Vieilledent2013}{2013}) associated with
human activities such as slash-and-burn agriculture and pasture (Scales
\protect\hyperlink{ref-Scales2011}{2011}). Tropical forests in
Madagascar also store a large amount of carbon (Vieilledent \emph{et
al.} \protect\hyperlink{ref-Vieilledent2016}{2016}) and high rates of
deforestation in Madagascar are responsible for large CO\(_2\) emissions
in the atmosphere (Achard \emph{et al.}
\protect\hyperlink{ref-Achard2014}{2014}). Deforestation threatens
species survival by directly reducing their available habitat (Tidd,
Pinder \& Ferguson \protect\hyperlink{ref-Tidd2001}{2001}; Brooks
\emph{et al.} \protect\hyperlink{ref-Brooks2002}{2002}). Forest
fragmentation can also lead to species extinction by isolating
populations from each other and creating forest patches too small to
maintain viable populations (Saunders, Hobbs \& Margules
\protect\hyperlink{ref-Saunders1991}{1991}). Fragmentation also
increases forest edge where ecological conditions (such as air
temperature, light intensity and air moisture) can be dramatically
modified, with consequences on the abundance and distribution of species
(Murcia \protect\hyperlink{ref-Murcia1995}{1995}). Forest fragmentation
can also have substantial effects on forest carbon storage capacity, as
carbon stocks are much lower at the forest edge than under a closed
canopy (Brinck \emph{et al.} \protect\hyperlink{ref-Brinck2017}{2017}).
Moreover, forest carbon stocks vary spatially due to climate or soil
factors (Saatchi \emph{et al.}
\protect\hyperlink{ref-Saatchi2011}{2011}; Vieilledent \emph{et al.}
\protect\hyperlink{ref-Vieilledent2016}{2016}). As a consequence,
accurate and spatially explicit maps of forest cover and forest cover
change are necessary to monitor biodiversity loss and carbon emissions
from deforestation and forest fragmentation, assess the efficiency of
present conservation strategies (Eklund \emph{et al.}
\protect\hyperlink{ref-Eklund2016}{2016}), and implement new strategies
for the future (Vieilledent, Grinand \& Vaudry
\protect\hyperlink{ref-Vieilledent2013}{2013}; Vieilledent \emph{et al.}
\protect\hyperlink{ref-Vieilledent2016}{2016}). Simple time-series of
forest cover estimates, such as those provided by the FAO Forest
Resource Assessment report (Keenan \emph{et al.}
\protect\hyperlink{ref-Keenan2015}{2015}) are not sufficient.

Unfortunately, accurate and exhaustive forest cover maps are not
available for Madagascar after year 2000. Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}) produced maps of forest cover
and forest cover changes over Madagascar for the years \emph{c.}~1953,
\emph{c.}~1973, 1990 and 2000. The \emph{c.}~1953 forest map was derived
from the visual interpretation of aerial photography at coarse scale
(1/1,000,000). Forest maps for the years \emph{c.}~1973, 1990, and 2000
were obtained from supervised classification of Landsat satellite images
at 60~m resolution (for the year 1973) or 30~m resolution (for years
1990 and 2000) and can be used to derive more accurate estimates of
forest cover (89.5\% accuracy reported for the forest/non-forest map of
year 2000). Nonetheless, maps provided by Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}) are not exhaustive (due to
the presence of clouds in the satellite imagery), e.g.~11 244~km2 are
mapped as unknown cover type for the year 2000. Using a similar
supervised classification approach as in Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}), more recent maps have been
produced for the periods 2000-2005-2010 by national institutions, with
the technical support of international environmental NGOs (MEFT, USAID,
and CI \protect\hyperlink{ref-MEFT2009}{2009}; ONE, DGF, FTM, MNP, and
CI \protect\hyperlink{ref-ONE2013}{2013}). Another set of recent forest
cover maps using an advanced statistical tool for classification, the
Random Forest classifier (Grinand \emph{et al.}
\protect\hyperlink{ref-Grinand2013}{2013}; Rakotomala \emph{et al.}
\protect\hyperlink{ref-Rakotomalala2015}{2015}), was produced for the
periods 2005-2010-2013 (ONE, DGF, MNP, WCS, and Etc Terra
\protect\hyperlink{ref-ONE2015}{2015}). However, these maps are either
too old to give recent estimates of deforestation (MEFT, USAID, and CI
\protect\hyperlink{ref-MEFT2009}{2009}; ONE, DGF, FTM, MNP, and CI
\protect\hyperlink{ref-ONE2013}{2013}), include large areas of missing
information due to images with high percentage of cloud cover (ONE, DGF,
FTM, MNP, and CI \protect\hyperlink{ref-ONE2013}{2013}), or show large
mis-classification in specific areas, especially in the dry and spiny
forest domain for which the spectral answer has a strong seasonal
behavior due to the deciduousness of such forests (overall accuracy is
lower than 0.8 for the dry and spiny forests for the maps produced by
ONE, DGF, MNP, WCS, and Etc Terra
(\protect\hyperlink{ref-ONE2015}{2015})). Moreover, the production of
such forest maps from a supervised classification approach requires
significant resources, especially regarding the image selection step
(required to minimize cloud cover) and the training step (visual
interpretation of a large number of polygons needed to train the
classification algorithm) (Rakotomala \emph{et al.}
\protect\hyperlink{ref-Rakotomalala2015}{2015}). Most of this work of
image selection and visual interpretation would need to be repeated to
produce new forest maps in the future using a similar approach.

Global forest or tree cover products have also been published recently
and can be tested at the national scale for Madagascar. Kim \emph{et
al.} (\protect\hyperlink{ref-Kim2014}{2014}) produced a global forest
cover change map from 1990 to 2000 (derived from Landsat imagery). This
product was updated to cover the period 1975-2005
(\url{http://glcf.umd.edu/data/landsatFCC/}) but forest cover maps after
2005 were not produced. Moreover, the approach used in Kim \emph{et al.}
(\protect\hyperlink{ref-Kim2014}{2014}) did not accurately map the
forests in the dry and spiny ecosystems of Madagascar (see Fig. 8 in Kim
\emph{et al.} (\protect\hyperlink{ref-Kim2014}{2014})). Hansen \emph{et
al.} (\protect\hyperlink{ref-Hansen2013}{2013}) mapped tree cover
percentage, annual tree cover loss and gain from 2000 to 2012 at global
scale at 30 m resolution. This product has since been updated and is now
available up to the year 2014 (Hansen \emph{et al.}
\protect\hyperlink{ref-Hansen2013}{2013}). To map forest cover from the
Hansen \emph{et al.} (\protect\hyperlink{ref-Hansen2013}{2013}) product,
a tree cover threshold must be selected (that defines forest cover).
Selecting such a threshold is not straightforward as the accuracy of the
global tree cover map strongly varies between forest types, and is
substantially lower for dry forests than for moist forests (Bastin
\emph{et al.} \protect\hyperlink{ref-Bastin2017}{2017}). Moreover, the
Hansen \emph{et al.} (\protect\hyperlink{ref-Hansen2013}{2013}) product
does not provide information on land-use. In particular the global tree
cover map does not separate tree plantations such as oil palm or
eucalyptus plantations from natural forests (Tropek \emph{et al.}
\protect\hyperlink{ref-Tropek2014}{2014}). Thus, the global tree cover
map from Hansen \emph{et al.} (\protect\hyperlink{ref-Hansen2013}{2013})
cannot be used alone to produce a map of forest cover (Tyukavina
\emph{et al.} \protect\hyperlink{ref-Tyukavina2017}{2017}).

In this study, we present a simple approach which combines the
historical forest maps from Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}) and more recent global
products from Hansen \emph{et al.}
(\protect\hyperlink{ref-Hansen2013}{2013}) to derive annual wall-to-wall
forest cover change maps over the period 2000-2014 for Madagascar. We
use the forest cover map provided by Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}) for the year 2000 (defining
the land-use) with the tree cover loss product provided by Hansen
\emph{et al.} (\protect\hyperlink{ref-Hansen2013}{2013}) that we apply
only inside forest areas identified by Harper \emph{et al.}
(\protect\hyperlink{ref-Harper2007}{2007}). Similar to the approach of
Harper \emph{et al.} (\protect\hyperlink{ref-Harper2007}{2007}), we also
assess trends in deforestation rates and forest fragmentation from
\emph{c.}~1953 to 2014. We finally discuss the possibility to extend our
approach to other tropical countries or repeat it in the future. We also
discuss how our results could help assess the effectiveness of current
conservation strategies, and implementation new conservation strategies
for the future in Madagascar.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}
